\documentclass[11pt,a4paper]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{geometry}
\geometry{margin=2.5cm}

\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage[authoryear]{natbib}

% --- Metadata --
\title{Spatial patterns of reported crime in Italy before, during and after the COVID-19 pandemic}
\author{Michele Brunelli}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This study examines the spatial patterns of reported crimes across Italian provinces over the period 2014--2023, with the aim of assessing whether the COVID-19 pandemic altered not only the intensity but also the geographic distribution of criminal activity. Crime data provided by Istat were normalised per 100,000 inhabitants and aggregated into three temporal phases: a pre-pandemic baseline (2014--2019), the acute pandemic period (2020--2021), and the post-pandemic recovery (2022--2023). The analysis was conducted at three nested geographic levels---macro-areas (NUTS-1), regions (NUTS-2), and provinces (NUTS-3)---using exploratory spatial analysis techniques, including global spatial autocorrelation (Moran's I) and local indicators of spatial association (LISA). Results reveal a clear dichotomy: contact-based crimes such as pickpocketing and residential burglary declined sharply, while digital offences including cybercrime and online fraud increased substantially. Spatial autocorrelation analysis shows that the pandemic fragmented the overall spatial structure of crime, though with heterogeneous effects across crime types. LISA cluster tracking identifies three distinct evolutionary trajectories: permanent geographic reorganisation for opportunity-dependent crimes, moderate spatial redistribution for digital offences, and high structural resilience for residential burglary. Sexual violence emerges as an anomalous case, displaying progressive spatial clustering intensification across all periods. A reproducible analysis pipeline and an interactive web application were developed to support the exploration of these patterns.
\end{abstract}

\section{Introduction}
\input{sections/introduction}

\section{Data}
\input{sections/data}

\section{Methodology}
\input{sections/methodology}

\section{Results}
\input{sections/results}

\section{Conclusions}
\input{sections/conclusions}

\bibliographystyle{apalike}
\bibliography{references}

\end{document}